The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-23 crelle-dictionary.tex
Lots of spacing problems. We'd better to this together, tomorrow  . . .

%%%%%%%%%%%%%%%%% PACKAGES %%%%%%%%%%%%%%%%%
% Encodings
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
% AMS packages
\usepackage{amsmath, amsthm, amssymb, todonotes}%, showkeys}
\usepackage[alphabetic]{amsrefs}
% Fonts
\usepackage{mathrsfs, yfonts}
%Geometry
\usepackage{geometry}
% Enumitem
%\usepackage{enumitem}
% TikZ
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,calc,matrix}
\usepackage{tikz-cd}
% Hyperrefs
\usepackage{hyperref}
\setcounter{section}{-1}
%%%%%%%%%%%%%%% THEOREM STYLES %%%%%%%%%%%%%%%
\theoremstyle{plain}
      \newtheorem{theorem}{Theorem}[section]
      \newtheorem{proposition}[theorem]{Proposition}
      \newtheorem{lemma}[theorem]{Lemma}
      \newtheorem{corollary}[theorem]{Corollary}

@@

We close the introduction with an overview of the structure of this paper.
After setting some basic notation in Section~\ref{sec:notation},
in Section~\ref{sec:category} we define the category of quasicharacter sheaves.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism $\TrFrob{G} : \QCiso{G} \to \Hom(G(\FF_q), \EEx)$.
We also define pullbacks of quasicharacter sheaves along morphisms of group schemes.  Applying pullbacks to the component group
sequence for $G$ yields the diagram
 \begin{equation} \label{eq:intro-snake}
  \begin{tikzcd}[row sep=20]
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \Hom(\pi_0(G)(\FF_q),\EEx) \rar & \Hom(G(\Fq),\EEx) \rar & \Hom(G^0(\FF_q),\EEx) \rar & 0,
  \end{tikzcd}
  \end{equation}
which is central to the proof of Theorem~\ref{thm:snake}.  Much of Sections~\ref{sec:etale}, \ref{sec:disc-isog} and \ref{sec:main}
are devoted to proving that the rows of \eqref{eq:intro-snake} are exact and determining the kernels and cokernels of the vertical maps.

@@

In Section~\ref{sec:main} we prove that the rows of \eqref{eq:intro-snake} are exact and
use the snake lemma to determine the kernel and cokernel of $\TrFrob{G}$ for general $G$.
We show that $\QCiso{G^0} \to \QCiso{G}$ is surjective in Section~\ref{ssec:restriction} and
that the remainder of the top row of \eqref{eq:intro-snake} is exact in Section~\ref{ssec:component}.
We put it together in Theorem~\ref{thm:snake} of Section~\ref{ssec:snake}, showing that $\TrFrob{G}$
is surjective with kernel $\Hh^2(\pi_0(\bG), \EEx)^{\Frob{}}$.  Having described the isomorphism classes of objects,
in Theorem~\ref{thm:autornaught} of Section~\ref{ssec:qcmor} we show that
\[
\Aut(\qcs{L}) \cong \Hom(\pi_0(\bG)_{\Weil{}}, \EEx)
\]
and that every morphism is either trivial or an isomorphism.  We conclude in Section~\ref{sec:basechange} by
giving functorial versions of the restriction and norm homomorphisms $\Hom(G(k'),\EEx) \to \Hom(G(k),\EEx)$ and $\Hom(G(k),\EEx) \to \Hom(G(k'),\EEx)$
and describing how quasicharacter sheaves behave under Weil restriction.

@@

The category $\QC(G)$ of quasicharacter sheaves on $G$ is a rigid monoidal category
\cite{etingof:09a}*{\S1.10} under the tensor product
$\qcs{L} \otimes \qcs{L'} \ceq (\gqcs{L}\otimes\gqcs{L'}, \mu\otimes\mu', \phi\otimes \phi')$,
with duals given by applying the sheaf hom functor
$\sheafHom(\ - \ ,\EE)$.
This rigid monoidal category structure for $\QC(G)$ gives the set $\QCiso{G}$
of isomorphism classes in $\QC(G)$ the structure of a group.

@@

In this section we introduce two tools which will help us understand isomorphism classes of objects in $\QC(G)$: the trace of Frobenius map $\QCiso{G} \to G(k)^*$ and the pullback functor $\QC(G) \to \QC(H)$ associated to a morphism $H \to G$.

@@

Now suppose $\qcs{L} = (\gqcs{L},\mu,\phi)$ is a quasicharacter sheaf on $G$.
The isomorphism $\mu : \bm^* \gqcs{L} \to \LxL$ together with
\ref{QC.3} guarantee
that the function $\trFrob{(\gqcs{L},\phi)} : G(\Fq)\to \EEx$ is a group homomorphism, which we will also denote by $\trFrob{\qcs{L}}$.  Moreover,
this homomorphism depends only on the isomorphism class of $\qcs{L}$, so we obtain a map
\begin{align*}
\TrFrob{G} : \QCiso{G} &\to G(\Fq)^* \\
\qcs{L} &\mapsto \trFrob{\qcs{L}}.
\end{align*}
Since tensor products on the stalks of $\qcs{L}$ induce pointwise multiplication on the trace of Frobenius, $\TrFrob{G}$ is a group homomorphism.  We devote the remainder of this section to
studying the behavior of the trace of Frobenius under morphisms of group schemes.  These simple but important
results on pullbacks and external products will be used in the proofs of many results in this paper.

@@

\begin{lemma}\label{lem:product}
If $G_1$ and $G_2$ are smooth commutative group schemes over $\Fq$ then the rule
$\boxtimes : (\qcs{L}_1,\qcs{L}_2)\to \qcs{L}_1\boxtimes\qcs{L}_2$ defines an equivalence of categories
\[
\QC(G_1)\times \QC(G_2) \to \QC(G_1\times G_2)
\]
such that
\[
\begin{tikzcd}[column sep=60]
\arrow{d}{\TrFrob{G_1} \times \TrFrob{G_2}} \QCiso{G_1}\times \QCiso{G_2} \arrow{r}{\boxtimes}
& \arrow{d}{\TrFrob{G_1\times G_2}} \QCiso{G_1\times G_2}\\
(G_1)(\Fq)^*\times (G_2)(\Fq)^* \arrow{r}{(\chi_1,\chi_2)\mapsto \chi_1\otimes \chi_2}  & (G_1\times G_2)(\Fq)^*
\end{tikzcd}
\]
commutes.
\end{lemma}
\begin{proof}
The only non-trivial part of the proof is to show that $\boxtimes : \QC(G_1)\times \QC(G_2) \to \QC(G_1\times G_2)$
is essentially surjective, so we will only address that point here.

>>>>

%%%%%%%%%%%%%%%%% PACKAGES %%%%%%%%%%%%%%%%%
% Encodings
%\usepackage[utf8]{inputenc}
%\usepackage[T1]{fontenc}
% AMS packages
%\usepackage{amsmath, amsthm, amssymb}
\usepackage{todonotes}%, showkeys}
\usepackage[alphabetic]{amsrefs}
% Fonts
\usepackage{mathrsfs, yfonts}
%Geometry
%\usepackage{geometry}
% Enumitem
%\usepackage{enumitem}
% TikZ
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,calc,matrix}
\usepackage{tikz-cd}
% Hyperrefs
\usepackage{hyperref}
\setcounter{section}{-1}
%%%%%%%%%%%%%%% THEOREM STYLES %%%%%%%%%%%%%%%
\theoremstyle{plain}
      \newtheorem{theorem}{Theorem}[section]
      \newtheorem{proposition}[theorem]{Proposition}
      \newtheorem{lemma}[theorem]{Lemma}
      \newtheorem{corollary}[theorem]{Corollary}

@@

We close the introduction with an overview of the structure of this paper.
After setting some basic notation in Section~\ref{sec:notation},
in Section~\ref{sec:category} we define the category of quasicharacter sheaves.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism $\TrFrob{G} : \QCiso{G} \to \Hom(G(\FF_q), \EEx)$.
We also define pullbacks of quasicharacter sheaves along morphisms of group schemes.  Applying pullbacks to the component group
sequence for $G$ yields the diagram
 \begin{equation} \label{eq:intro-snake}
  \begin{tikzcd}[row sep=20, column sep=20]
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \Hom(\pi_0(G)(\FF_q),\EEx) \rar & \Hom(G(\Fq),\EEx) \rar & \Hom(G^0(\FF_q),\EEx) \rar & 0,
  \end{tikzcd}
  \end{equation}
which is central to the proof of Theorem~\ref{thm:snake}.  Much of Sections~\ref{sec:etale}, \ref{sec:disc-isog} and \ref{sec:main}
are devoted to proving that the rows of \eqref{eq:intro-snake} are exact and determining the kernels and cokernels of the vertical maps.

@@

In Section~\ref{sec:main} we prove that the rows of \eqref{eq:intro-snake} are exact and
use the snake lemma to determine the kernel and cokernel of $\TrFrob{G}$ for general $G$.
We show that $\QCiso{G^0} \to \QCiso{G}$ is surjective in Section~\ref{ssec:restriction} and
that the remainder of the top row of \eqref{eq:intro-snake} is exact in Section~\ref{ssec:component}.
We put it together in Theorem~\ref{thm:snake} of Section~\ref{ssec:snake}, showing that $\TrFrob{G}$
is surjective with kernel $\Hh^2(\pi_0(\bG), \EEx)^{\Weil{}}$.  Having described the isomorphism classes of objects,
in Theorem~\ref{thm:autornaught} of Section~\ref{ssec:qcmor} we show that
\[
\Aut(\qcs{L}) \cong \Hom(\pi_0(\bG)_{\Weil{}}, \EEx)
\]
and that every morphism is either trivial or an isomorphism.  We conclude in Section~\ref{sec:basechange} by
giving functorial versions of the restriction and norm homomorphisms 
of character groups
%$\Hom(G(k'),\EEx) \to \Hom(G(k),\EEx)$ and $\Hom(G(k),\EEx) \to \Hom(G(k'),\EEx)$
and describing how quasicharacter sheaves behave under Weil restriction.

@@

Category $\QC(G)$ is a rigid monoidal category
\cite{etingof:09a}*{\S1.10} under the tensor product
$\qcs{L} \otimes \qcs{L'} \ceq (\gqcs{L}\otimes\gqcs{L'}, \mu\otimes\mu', \phi\otimes \phi')$,
with duals given by applying the sheaf hom functor
$\sheafHom(\ - \ ,\EE)$.
This rigid monoidal category structure for $\QC(G)$ gives the set $\QCiso{G}$
of isomorphism classes in $\QC(G)$ the structure of a group.

@@

In this section we introduce two tools which will help us understand isomorphism classes of objects in $\QC(G)$: the map $\QCiso{G} \to G(k)^*$ given by trace of Frobenius and the pullback functor $\QC(G) \to \QC(H)$ associated to a morphism $H \to G$.

@@

Now suppose $\qcs{L} = (\gqcs{L},\mu,\phi)$ is a quasicharacter sheaf on $G$.
The isomorphism $\mu : \bm^* \gqcs{L} \to \gqcs{L} \boxtimes\gqcs{L} $ together with
\ref{QC.3} guarantee
that the function $\trFrob{(\gqcs{L},\phi)} : G(\Fq)\to \EEx$ is a group homomorphism, which we will also denote by $\trFrob{\qcs{L}}$.  Moreover,
this homomorphism depends only on the isomorphism class of $\qcs{L}$, so we obtain a map
\begin{align*}
\TrFrob{G} : \QCiso{G} &\to G(\Fq)^* \\
\qcs{L} &\mapsto \trFrob{\qcs{L}}.
\end{align*}
Since tensor products on the stalks of $\qcs{L}$ induce pointwise multiplication on the trace of Frobenius, $\TrFrob{G}$ is a group homomorphism.  We devote the remainder of this section to
studying the behavior of the trace of Frobenius under morphisms of group schemes.  These simple but important
results on pullbacks and external products will be used in the proofs of many results in this paper.

@@

\begin{lemma}\label{lem:product}
If $G_1$ and $G_2$ are smooth commutative group schemes over $\Fq$ then the rule
$\boxtimes : (\qcs{L}_1,\qcs{L}_2)\to \qcs{L}_1\boxtimes\qcs{L}_2$ defines an equivalence of categories
\[
\QC(G_1)\times \QC(G_2) \to \QC(G_1\times G_2)
\]
such that
\[
\begin{tikzcd}[column sep=60]
\arrow{d}{\TrFrob{G_1} \times \TrFrob{G_2}} \QCiso{G_1}\times \QCiso{G_2} \arrow{r}{\boxtimes}
& \arrow{d}{\TrFrob{G_1\times G_2}} \QCiso{G_1\times G_2}\\
(G_1)(\Fq)^*\times (G_2)(\Fq)^* \arrow{r}{(\chi_1,\chi_2)\mapsto \chi_1\otimes \chi_2}  & (G_1\times G_2)(\Fq)^*
\end{tikzcd}
\]
commutes.
\end{lemma}
\begin{proof}
The only non-trivial part of the proof is to show that 
\[\boxtimes
 : \QC(G_1)\times \QC(G_2) \to \QC(G_1\times G_2)
 \]
is essentially surjective, so we will only address that point here.
